\documentclass[a4paper]{article}

\usepackage[pages=all, color=black, position={current page.south}, placement=bottom, scale=1, opacity=1, vshift=5mm]{background}
%\SetBgContents{
%	\tt This work is shared under a \href{https://creativecommons.org/licenses/by-sa/4.0/}{CC BY-SA 4.0 license} unless otherwise noted
%}      % copyright

\usepackage{physics}
\usepackage[margin=1in]{geometry} % full-width

% AMS Packages
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}

% Unicode
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\hypersetup{
	unicode,
%	colorlinks,
%	breaklinks,
%	urlcolor=cyan, 
%	linkcolor=blue, 
	pdfauthor={Author One, Author Two, Author Three},
	pdftitle={A simple article template},
	pdfsubject={A simple article template},
	pdfkeywords={article, template, simple},
	pdfproducer={LaTeX},
	pdfcreator={pdflatex}
}

% Natbib
\usepackage[sort&compress,numbers,square]{natbib}
\bibliographystyle{mplainnat}

% Theorem, Lemma, etc
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{claim}{Claim}[theorem]
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{fact}[theorem]{Fact}
\newtheorem{hypothesis}[theorem]{Hypothesis}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{criterion}[theorem]{Criterion}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{principle}[theorem]{Principle}

\usepackage{graphicx, color}
\graphicspath{{fig/}}

%\usepackage[linesnumbered,ruled,vlined,commentsnumbered]{algorithm2e} % use algorithm2e for typesetting algorithms
\usepackage{algorithm, algpseudocode} % use algorithm and algorithmicx for typesetting algorithms
\usepackage{mathrsfs} % for \mathscr command

% Author info
\title{Andy's math/science background information}
\author{J. Andrew Fingerhut (\texttt{andy.fingerhut@gmail.com})}

\date{
        June 1, 2025
%	\today
}

\newcommand{\ihat}{\textbf{i}}
\newcommand{\jhat}{\textbf{j}}
\newcommand{\khat}{\textbf{k}}
\newcommand{\vect}[1]{\textbf{#1}}
\newcommand{\hatvec}[1]{\hat{\textbf{#1}}}

\newcommand{\reals}{\mathbb{R}}
\newcommand{\del}{\nabla}
%\newcommand{\grad}{grad}

\begin{document}
\maketitle

%\begin{abstract}
%  todo: abstract here
%\end{abstract}

\tableofcontents

\section{Introduction}
\label{sec:intro}

This document is a place to write up notes on background in math that
help in learning science.


\section{Notation}
\label{sec:notation}

$\reals^3$ - The set of all points in 3-dimensional space,
i.e. where each point is specified by a sequence of 3 real-valued
coordinates.

$\del$ - Called ``nabla'', and often called ``del''.  Some history of
it can be found on the Nabla Symbol page~\cite{NablaSymbol}.

$\grad f$ - Gradient of a scalar function $f: \reals^3 \rightarrow
\reals$.

$\text{grad} f$ - another way to write the gradient of $f$, $\grad f$

TODO: Any 3Blue1Brown-quality YouTube videos showing examples and
definition of gradient, divergence, and curl?  If so, give links to
them below in appropriate sections.


\subsection{Maxwell's Equations}
\label{sec:maxwellseqns}

In partial differentiatal form:

\begin{align}
  \grad \cdot \vect{E} & = \frac{\rho}{\epsilon_0} & & \text{Gauss's Law} \label{eq:gaussE} \\
  \grad \cdot \vect{B} & = 0 & & \text{Gauss's Law for magnetism} \label{eq:gaussB} \\
  \grad \times \vect{E} & = - \frac{\partial \vect{B}}{\partial t} & & \text{Maxwell-Faraday Equation (Faraday's law of induction)} \label{eq:maxfar} \\
  \grad \times \vect{B} & = \mu_0 ( \vect{J} + \epsilon_0 \frac{\partial \vect{E}}{\partial t} ) & & \text{Amp\`{e}re-Maxwell law} \label{eq:ampmax}
\end{align}
where:
\begin{itemize}
  \item $\vect{E}$ is the electric field
  \item $\vect{B}$ is the magnetic field
  \item $\rho$ is the electric charge density
  \item $\vect{J}$ is the current density
  \item $\epsilon_0$ is the vacuum permittivity
  \item $\mu_0$ is the vacuum permeability
\end{itemize}

Statement of the equations in English:

Gaus's Law (equation~\eqref{eq:gaussE}) says that electric fields are
``produced'' radially outward from volumes with positive electric
charge, and ``consumed'' radially into volumes with negative electric
charge.

TODO: Is this a strict generalization of Coulomb's Law?  In what ways?
One way is that Gauss's Law gives the electric field generated by a
distribution of charge, whereas Coulomb's Law gives the force between
two point charges.  So applying to a continuous distribution of charge
is one generalization, and what electric field is created by that
distribution of charge is another (as long as you then combine Gauss's
Law with the Lorentz force law in order to calculate the forces).

TODO: It seems that Gauss's Law, if you take $\rho$ and $\vect{E}$ as
functions of time, gives an electric field that changes
instantaneously everywhere when the charge distribution $\rho$
changes.  There does not appear to be anything related to retarded
positions of charge in that equation anywhere.  How does the finite
speed of electric field propagation enter into Maxwell's equations?

Doing a Google search on the question ``is Gauss's law consistent with
relativistic effects'' gives answers that it {\emi is} consistent with
special relativity, including the following paragraphs:

\begin{itemize}
\item
Maxwell's equations are relativistic: Gauss's law is one of Maxwell's
equations.  These equations, which describe classical
electromagnetism, are inherently Lorentz invariant.  This means they
hold true in all inertial reference frames, consistent with the
principle of relativity, which states that the laws of physics are the
same for all observers in uniform motion relative to one another.
\item
Lorentz covariance: Maxwell's equations, and thus Gauss's law, can be
written in a covariant form that is explicitly consistent with Lorentz
transformations.  These transformations describe how physical
quantities change when moving from one inertial frame to another,
incorporating relativistic effects like time dilation and length
contraction.
\end{itemize}


\section{Gradient}
\label{sec:gradient}

Gradient has been generalized to many coordinate systems other than
the 3-dimensional Cartesian coordinate system $\mathbb{R}^3$, but I
will focus on $\mathbb{R}^3$.  The Wikipedia page on
Gradient~\cite{Gradient} is not too bad for me, as long as I skim over
the parts that generalize it to other coordinate systems.

Griffiths~\cite{Griffiths1998} Section 1.2.2 ``Gradient'' is good at
giving the definition and some useful examples and properties of the
gradient.  He defines $\grad T$ this way, where $\hatvec{x}$,
$\hatvec{y}$, and $\hatvec{z}$ are the unit vectors in the direction
of the three coordinate axes:
\begin{align*}
  \grad T & \equiv \frac{\partial T}{\partial x} \hatvec{x}
                 + \frac{\partial T}{\partial y} \hatvec{y}
                 + \frac{\partial T}{\partial z} \hatvec{z}
\end{align*}
where:
\begin{itemize}
  \item $T : \reals^3 \rightarrow \reals$ is a scalar function that is
    continuous and differentiable
\end{itemize}
Some properties of the gradient:
\begin{itemize}
  \item $\grad T(x, y, z)$ is the gradient evaluated at a position
    given by $(x, y, z)$.  The vector points in a direction that
    function $T$ increases most quickly.
  \item The function $T$ can be approximated at points $\vect{r}$ near
    $\vect{r}_0 = (x, y, z)$ by the linear function $T(\vect{r}_0) +
    (\vect{r} - \vect{r}_0) \cdot \grad T(\vect{r}_0)$
  \item The instantaneous rate of change of $T$ in direction
    $\vect{u}$ (a unit vector) from $\vect{r}_0$ is $\vect{u} \cdot
    \grad T(\vect{r}_0)$.  Note that it is always 0 in a direction
    perpendicular to $\grad T(\vect{r}_0)$, and the negative of the
    magnitude of $\grad T(\vect{r}_0)$ in the opposite direction.
  \item Consider a ``level set'', i.e. a surface defined by all of the
    points $\vect{r}$ where $T(\vect{r})=c$ for some constant $c$.
    Then $\grad T$ evaluated at any point on that surface, is normal
    to the surface.
\end{itemize}


%\newpage
\bibliography{refs}

%\appendix

%\section{Omitted Proof in Section~\ref{sec:examples}}
%\label{app:1}

	
\end{document}
